Learning Discrete-valued Bayesian Networks from Mixed Data

### 2015-11-19 BN_discretizer_paper.tex
Correct things

\begin{algorithm}
  \caption{Discretization of one continuous variable in a Bayesian network}
  \label{alg:disc_one}
  \begin{algorithmic}[5]
    \Function{DiscretizeOne}{$\cX$, $D$, $G$}
      \State $n \leftarrow$ the number of data instances
      \State $m \leftarrow$ the number of unique values in $D_{\cX}$
      \State $H \leftarrow$ an $N \times N$ matrix such that $H[u,v] = h(u,v)$ (see Algorithm~\ref{alg:h} in Appendix)
      \State $s_{1:m} \leftarrow$ an increasing sequence of integers defined in Section \ref{subsec:disc_policy}
      \State $L \leftarrow$ the largest cardinality over all discrete variables in the Markov blanket of $X$
      \State $S[u] \leftarrow$ the optimal objective value of subproblems from instance $1$ to $s_u$
      \State $\Lambda[u] \leftarrow$ the optimal discretization for subproblems from instance $1$ to $s_u$
      \State $W[u]  \leftarrow - \ln\brock{1 - {\exp\paren{- L \cdot{ {{\cx_{s_u+1} - \cx_{s_u}}\over{\cx_n - \cx_1}}}}}}$ for $u = 1,2, \ldots,m-1$ and $W[n] \leftarrow 0$
      \For {$v \leftarrow 1$ to $m$}
        \If {$v = 1$}
          \State $S[v] \leftarrow H \paren{1,s_v} + L[v]$
          \State $\Lambda[v] \leftarrow \{ ({\cx_{s_v} + \cx_{s_v+1}}) / 2\}$
        \Else
          \State $s \leftarrow \infty$
          \State $\text{DiscEdge} \leftarrow \infty$
        \For {$u \leftarrow 1$ to $v$}
        \State $s' \leftarrow S[u] + H \paren{s_u+1, s_v} +  {L \cdot {\frac{\cx_{s_v} - \cx_{s_u + 1}}{\cx_n - \cx_1}}} + W[v]$
        \If {$s' < s$}
          \State $s \leftarrow s'$
          \State $\text{DiscEdge} \leftarrow ({x_{s_u} + x_{s_u+1}}) / 2$
        \EndIf
        \EndFor
          \State $S[v] \leftarrow s$
          \State $\Lambda[v] \leftarrow M[u] \cup \{ DiscEdge \}$
        \EndIf
      \EndFor
      \State \Return $\Lambda[m]$
    \EndFunction
  \end{algorithmic}
\end{algorithm}

@@

Table \ref{table:auto_disc_table_1} shows the positions of discretization edges, and log-likelihood of 10-fold cross validation by three discretization methods. For the Bayesian network in Fig. \ref{fig:auto_graph_1}, the approximation \ref{eq:opt_prob_approx} only reduces runtime and has no influence on discretization result. MDL principle discretization works not well, since it discretizes each continuous variable by only one interval. The reason that the MDL method discretizes variables into too few number of intervals is discussed in Section \ref{subsec:discuss_exp}.

@@

The Bayesian network in Figure \ref{fig:wine_graph_1} is obtained by equal-width discretizing each continuous variable into 3 intervals and running K2 a thousand times. Table \ref{table:wine_disc_table_1} shows a comparison between the optimal Bayesian approach and the MDL principle discretization method.
The discretization result obtained by the optimal Bayesian approach is better than the MDL discretization method in terms of likelihood. The heuristic method discretizes variable $G$, $H$ and $J$ into one less interval than the optimal Bayesian approach, and has one more number of intervals on variable $\Lambda$.
Although it returns different numbers of intervals and different positions of discretization edges with the optimal Bayesian approach, it still has a good discretization result in terms of likelihood. Furthermore, if running time is the main concern, the heuristic method can be considered a better approach in this case.

@@

Figure \ref{fig:wine_exp2_distr} shows the discretization result on variable $E$ and $K$. The discretization edge at \num{17.9} on variable $E$ also appers in Figure \ref{fig:wine_exp1_distr}. This indicates that some discretization edges are robust against the network structure. Furthermore, in Figure \ref{fig:wine_exp2_distr}, the discretization edge at \num{23.5} on $E$ is lost. This is due to the number of parents of $E$ in Figure \ref{fig:wine_graph_2} is twice as in \ref{fig:wine_graph_1}. Usually, if a variable has more parents, then less number of intervals after discretization it has, since the information from parents are adulterated together.

@@

Figure~\ref{fig:housing_exp1_distr_3_5} and Figure~\ref{fig:housing_exp1_distr_8_5} show the discretization result of the optimal Bayesian approach on the network in Figure~\ref{fig:housing_graph_1}.
In Figure~\ref{fig:housing_exp1_distr_3_5}, the scatter points, which stands for the raw data, has been jittered, since there are many repeated values in the raw data on the variable pair $C$ and $E$. These repeating values cause the high number of intervals for both $C$ and $E$, because in some sense, each repeated-value point form a single discretized area. On the other hand, there is no such over repeated values of the raw data on variable $H$. Thus, the numbers of intervals after discretization for variable $H$ remains normal, as what we have seen in previous datasets.

@@

Figure~\ref{fig:housing_exp2_distr_3_5} and Figure~\ref{fig:housing_exp2_distr_8_5} show the discretization result on variables of the network in Figure~\ref{fig:housing_graph_2}.
Again, variable $C$ and $E$ have many discretization edges due to the repeated values.
Although the number of intervals after discretization on $H$ is not large, it captures the distribution of the raw data along with the discretization edges on $E$.

@@

\begin{figure}[ht]
\centering
    % \begin{tabular}{cc}
      \input{graph_housing_exp2_distr_8_5}
    % \end{tabular}
  \caption{The discretization result on variable $H$ and $E$ of the network in Figure~\ref{fig:housing_graph_2}. Although the number of discretization edges on $H$ is not as large as $E$'s, they still match the distribution of the raw data very well.}
  \label{fig:housing_exp2_distr_8_5}
\end{figure}

>>>>

\begin{algorithm}
  \caption{Discretization of one continuous variable in a Bayesian network}
  \label{alg:disc_one}
  \begin{algorithmic}[5]
    \Function{DiscretizeOne}{$\cX$, $D$, $G$}
      \State $n \leftarrow$ the number of data instances
      \State $m \leftarrow$ the number of unique values in $D_{\cX}$
      \State $H \leftarrow$ an $N \times N$ matrix such that $H[u,v] = h(u,v)$ (see Algorithm~\ref{alg:h} in Appendix)
      \State $s_{1:m} \leftarrow$ an increasing sequence of integers defined in Section \ref{subsec:disc_policy}
      \State $L \leftarrow$ the largest cardinality over all discrete variables in the Markov blanket of $X$
      \State $S[u] \leftarrow$ the optimal objective value of subproblems from instance $1$ to $s_u$
      \State $\Lambda[u] \leftarrow$ the optimal discretization for subproblems from instance $1$ to $s_u$
      \State $W[u]  \leftarrow - \ln\brock{1 - {\exp\paren{- L \cdot{ {{\cx_{s_u+1} - \cx_{s_u}}\over{\cx_n - \cx_1}}}}}}$ for $u = 1,2, \ldots,m-1$ and $W[n] \leftarrow 0$
      \For {$v \leftarrow 1$ to $m$}
        \If {$v = 1$}
          \State $S[v] \leftarrow H \paren{1,s_v} + L[v]$
          \State $\Lambda[v] \leftarrow \{ ({\cx_{s_v} + \cx_{s_v+1}}) / 2\}$
        \Else
          \State $s \leftarrow \infty$
          \State $\text{DiscEdge} \leftarrow \infty$
        \For {$u \leftarrow 1$ to $v$}
        \State $s' \leftarrow S[u] + H \paren{s_u+1, s_v} +  {L \cdot {\frac{\cx_{s_v} - \cx_{s_u + 1}}{\cx_n - \cx_1}}} + W[v]$
        \If {$s' < s$}
          \State $s \leftarrow s'$
          \State $\text{DiscEdge} \leftarrow ({x_{s_u} + x_{s_u+1}}) / 2$
        \EndIf
        \EndFor
          \State $S[v] \leftarrow s$
          \State $\Lambda[v] \leftarrow \Lambda[u] \cup \{ DiscEdge \}$
        \EndIf
      \EndFor
      \State \Return $\Lambda[m]$
    \EndFunction
  \end{algorithmic}
\end{algorithm}

@@

Table \ref{table:auto_disc_table_1} shows the positions of discretization edges, and log-likelihood of 10-fold cross validation by three discretization methods. For the Bayesian network in Fig. \ref{fig:auto_graph_1}, the approximation Equation \ref{eq:opt_prob_approx} only reduces runtime and has no influence on discretization result. MDL principle discretization works not well, since it discretizes each continuous variable by only one interval. The reason that the MDL method discretizes variables into too few number of intervals is discussed in Section \ref{subsec:discuss_exp}.

@@

The Bayesian network in Figure \ref{fig:wine_graph_1} is obtained by equal-width discretizing each continuous variable into 3 intervals and running K2 a thousand times. Table \ref{table:wine_disc_table_1} shows a comparison between the optimal Bayesian approach and the MDL principle discretization method.
The discretization result obtained by the optimal Bayesian approach is better than the MDL discretization method in terms of likelihood. The heuristic method discretizes variable $G$, $H$ and $J$ into one less interval than the optimal Bayesian approach, and has one more number of intervals on variable $M$.
Although it returns different numbers of intervals and different positions of discretization edges with the optimal Bayesian approach, it still has a good discretization result in terms of likelihood. Furthermore, if running time is the main concern, the heuristic method can be considered a better approach in this case.

@@

Figure \ref{fig:wine_exp2_distr} shows the discretization result on variable $E$ and $K$. The discretization edge at \num{17.9} on variable $E$ also appers in Figure \ref{fig:wine_exp1_distr}. This indicates that some discretization edges are robust against the network structure. Furthermore, in Figure \ref{fig:wine_exp2_distr}, the discretization edge at \num{23.5} on $E$ is lost. This is due to the number of parents of $E$ in Figure \ref{fig:wine_graph_2} is twice as in \ref{fig:wine_graph_1}. Usually, if a variable has more parents, then less number of intervals after discretization it has, since the information from parents are mixed together.

@@

Figure~\ref{fig:housing_exp1_distr_3_5} and Figure~\ref{fig:housing_exp1_distr_8_5} show the discretization result of the optimal Bayesian approach on the network in Figure~\ref{fig:housing_graph_1}.
In Figure~\ref{fig:housing_exp1_distr_3_5}, the scatter points, which stands for the raw data, has been jittered, since there are many repeated values in the raw data on the variable pair $C$ and $E$. These repeating values cause the high number of intervals for both $C$ and $E$, because in some sense, each repeated-value point form a single discretized area. On the other hand, the data instances on variable $H$ is less repeated. Thus, some resulting discretized areas cover larger space, instead of just covering a repeated value.

@@

Figure~\ref{fig:housing_exp2_distr_3_5} and Figure~\ref{fig:housing_exp2_distr_8_5} show the discretization result on variables of the network in Figure~\ref{fig:housing_graph_2}.
Again, variable $C$ and $E$ have many discretization edges due to the repeated values.
Although the number of intervals after discretization on $H$ is less the number in Figure \ref{fig:housing_exp1_distr_8_5}, it still captures the distribution of the raw data along with the discretization edges on $E$.

@@

\begin{figure}[ht]
\centering
    % \begin{tabular}{cc}
      \input{graph_housing_exp2_distr_8_5}
    % \end{tabular}
  \caption{The discretization result on variable $H$ and $E$ of the network in Figure~\ref{fig:housing_graph_2}. Although the number of discretization edges on $H$ is less than the number in Figure \ref{fig:housing_exp1_distr_8_5}, it still matches the distribution of the raw data well.}
  \label{fig:housing_exp2_distr_8_5}
\end{figure}
